Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-09 manuscript/manuscript.tex
edits

One concern is that these covariances reflect the localized impact of a few
large-effect loci rather than selection on a polygenic trait. Since our
covariances are essentially averages over loci, the covariance estimate can be
strongly affected by a single outlier region. To test whether large outlier
regions drive the genome-wide signal we see in the \textcite{Barghi2019-qy}
data, we calculate the covariances in 100kb windows along the genome (we refer
to these as windowed covariances throughout) and take the median windowed
covariance, and trimmed-mean windowed covariance, as a measure of the
genome-wide covariance robust to large-effect loci. These robust estimates
(Supplementary Table \ref{supp:table-trimmed-mean} and Supplementary Figure
\ref{suppfig:barghi-trimmed-mean}) confirm the patterns we see using the mean
covariance, confirming that genomic temporal covariances are non-zero due to
the impact of selection \vb{acting across} many genomic windows.

@@

We've calculated the convergent correlation for all rows of the replicate
covariance matrices. Like temporal covariances, we visualize these through time
(Figure \ref{fig:figure-2}A), with each line representing the convergent
correlation from a particular reference generation $s$ as it varies with $t$
(shown on the x-axis). In other words, each of the colored lines corresponds to
the like-colored row of the convergence correlation matrix (upper left in
Figure \ref{fig:figure-2}A). We find these convergent covariances are
relatively weak, and decay very quickly from an initial value of about 0.1
(95\% block bootstrap confidence intervals $[0.094, 0.11]$) to around 0.01
(95\% CIs [0.0087, 0.015]) within 20 generations. This suggests that while
a reasonable fraction of the initial response is shared over the replicates,
this is followed by a rapid decay, a result consistent with the primary
finding of the original \textcite{Barghi2019-qy} study: that alternative loci
contribute to longer term adaptation \vb{across} the different replicates.

@@

Next, we reanalyzed the Longshanks selection experiment, which selected for
longer tibiae length relative to body size in mice, leading to a response to
selection of about 5 standard deviations over the course of twenty generations
\parencite{Marchini2014-de,Castro2019-uk}. This study includes two independent
selection lines, Longshanks 1 and 2 (LS1 and LS2), and an unselected control
line (Ctrl).  Consequently, this selection experiment offers a useful control
to test our between-replicate covariances: we expect to see positive
between-replicate covariance in the comparison between the two Longshanks
selection lines, but not between the two pairwise comparisons between the
control line and \vb{each of} the two Longshanks lines. We find that this is
the case (gray confidence intervals in Figure \ref{fig:figure-2} C), with the
two Longshanks comparisons to the control line not being significantly
different from zero, while the comparison between the two Longshanks lines is
statistically significantly different from zero (CIs [0.0129, 0.0400]).

One finding in the Longshanks study \vb{was that two} major-effect loci showed
parallel frequency shifts between the two selection lines: a region harboring
the gene Nkx3-2 known to be involved in limb development, and another region
containing six other candidate genes. We were curious to what extent our
genome-wide covariances were being driven by these two outlier large-effect
loci, so we excluded them from the analysis. Since we do not know the extent to
which linkage disequilibrium around these large-effect loci affects neighboring
loci, we took the conservative precaution of excluding the entire chromosomes
these loci reside on (chromosomes 5 and 10), and re-calculating the temporal
covariances. We find excluding these large effect loci has little impact on the
confidence intervals (blue confidence intervals in Figure \ref{fig:figure-2} C),
indicating that these across-replicate covariances are indeed driven by a large
number of loci. This is consistent with a signal of selection on a polygenic
trait driving genome-wide change, although we note that large-effect loci can
contribute to the indirect change at unlinked loci
\parencite{Robertson1961-ho,Santiago1995-hx}.

The presence of an unselected control line provides an alternative way to
partition the effects of linked selection and genetic drift: we can compare the
total variance in allele frequency change of the control line (which excludes
the effect of artificial selection on allele frequencies) to the total variance
in frequency change of the Longshanks selection lines. We can partition the
allele frequency change between the two timepoints (20 generations apart) for a
Longshanks line as \vb{$\Delta p_{t,\mathrm{LS1}} = \Delta_{_{D}}
  p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} + \Delta_{_S}
  p_{t,\mathrm{LS}}$} where these terms are the decomposition in the allele
  frequency change due to drift in \vb{Longshanks replicate 1 ($\Delta_{_D}
    p_{t,\mathrm{LS1}}$), selection unique to the LS1 replicate ($\Delta_{_U}
    p_{t,\mathrm{LS1}}$), and selection response shared between the two
    Longshanks replicates ($\Delta_{_S} p_{t,\mathrm{LS}}$) respectively (and
  similarly for the Longshanks two line, LS2)}. By construction we will assume
  that each of these terms are uncorrelated within replicates, and that only
  the shared term covaries between the replicates.  Assuming that we can
  approximate the contribution of genetic drift in the LS lines as the variance
  in allele frequency change observed in the control, i.e.  \vb{$\var(\Delta
    p_{t,\mathrm{Ctrl}}) = \var(\Delta_{_{D}}
  p_{t,\mathrm{LS2}})=\var(\Delta_{_{D}} p_{t,\mathrm{LS2}})$}, then we can
  estimate the increase in variance in allele frequency change due to selection
  as $(\var(\Delta p_{t,\mathrm{LS1}}) + \var(\Delta p_{t,\mathrm{LS2}}))/2 -
  \var(\Delta p_{t,\mathrm{Ctrl}})$ and the shared effect of selection across
  selected lines as $\cov(\Delta p_{t,\mathrm{LS1}}, \Delta
  p_{t,\mathrm{LS2}})$ (see Supplementary Material Section
  \ref{supp:replicate-g} for more details).  We estimate at least 32\% (95\% CI
  $[21\%, 48\%]$) of the variance in allele frequency change is driven by the
  effects of selection, of which 14\% (95\% CI $[3\%, 33\%]$) is estimated to
  be unique to a selection line, and 17\% (95\% CI $[9\%, 23\%]$) is the effect
  of shared selection between the two Longshanks selection lines.

@@

Finally, we observed that in the longest study  we analyzed
\parencite{Barghi2019-qy}, some genome-wide temporal covariances become
negative at future timepoints (see the first two rows in Figure
\ref{fig:figure-1} A). This shows that alleles that were on average going up
initially are later going down in frequency, i.e. that the average direction of
selection experienced by alleles has flipped. This must reflect either a change
in the environment or the genetic background, due to epistatic relationships
among alleles altered by frequency changes or recombination breaking up
selective alleles.  Such reversals of selective dynamics could be occurring at
other timepoints but the signal of a change in the direction of selection at
particular loci may be washed out when we calculate our genome-wide average
temporal covariances. To address this limitation, we calculated the
distribution of the temporal covariances over \vb{100kb windowed covariances}
(Figure \ref{fig:figure-3} shows these distributions pooling across all
replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance estimate of each genomic window will be noisy, due to sampling and
genetic drift, and the neutral distribution of the covariance is complicated
due to linkage disequilibria (which can occur over long physical distances in
E\&R and selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}). To
address this, we have developed a permutation-based procedure that constructs
an empirical null distribution by randomly flipping the signs of the allele
frequency changes per-genomic window. This destroys the systematic covariances
created by linked selection and creates a sampling distribution of the
covariances spuriously created by neutral genetic drift while preserving the
complex dependencies between adjacent loci created by linkage disequilibrium.
This empirical neutral null distribution is conservative in the sense that the
variances of the covariances are wider than expected under drift alone as they
include the effect of selection on the allele frequency change within a
time-interval, just not between time-intervals. We see (Figure
\ref{fig:figure-3} A and B) that windowed temporal covariances between close
timepoints are skewed positive (a heavy right tail), while between more distant
timepoints these windowed temporal covariances tend to shift to become more
negative (a heavy left tail).  We quantified the degree to which the left and
right tails are inflated compared to the null distribution as a function of
time, and see excesses in both tails in Figure \ref{fig:figure-3} C. This
finding is also robust to sign-permuting allele frequency changes on a
chromosome-level, the longest extent that gametic linkage disequilibria can
extend (Supplementary Figure \ref{suppfig:barghi-tailprobs-seqid}). We see a
striking pattern that the windowed covariances not only decay towards zero, but
in fact become negative through time, consistent with many regions in the
genome having had a reversed fitness effect at later timepoints.

@@

These estimates of the contribution of selection could be refined by using
patterns of LD and recombination which would allow us to more fully
parameterize a linked-selection model of temporal allele frequency change
\parencite{Buffalo2019-io}. The basic prediction is that regions of higher
linkage disequilibrium and lower recombination should have greater temporal
autocovariance than regions with lower LD and higher recombination. However,
one limitation of these pooled sequence datasets is that none of the studies we
reanalyzed estimated linkage disequilibria data for the evolved populations.
\vb{While} there are LD data for a natural population of \emph{D. simulans}
\parencite{Signor2018-wg,Howie2018-ay},  we did not find a relationship between
temporal covariance and LD.  We believe this is driven by the idiosyncratic
nature of LD in evolve-and-resequence populations, which often extends over
large genomic distances \parencite{Nuzhdin2013-gf,Kelly2019-dc}. Future studies
complete with LD data and recombination maps would allow one to disentangle the
influence of closely linked sites from more distant sites in causing temporal
autocovariance, and allow the fitting of more parametric models to estimate
population parameters such as the additive genetic variation for fitness
directly from temporal genomic data alone \parencite{Buffalo2019-io}.

Our primary focus here has been on evolution in laboratory populations. It is
unclear whether we should expect a similar impact of selection in natural
populations. In some of these experiments, selection pressures may have been
stronger or more sustained that in natural populations
\vb{\parencite{Hendry1999-zu,Hairston2005-ga}}. Conversely, these lab
populations were maintained at very small effective population sizes, estimated
at 300, 450, and 45 for the \textcite{Barghi2019-qy}, \textcite{Kelly2019-dc},
and \textcite{Castro2019-uk} studies respectively, which will amplify the role
of genetic drift. The advantage of lab experiments is that they are closed
populations, in natural populations temporal covariance could also arise from
the systematic migration of alleles from differentiated populations.
\vb{Adapting} these methods to natural populations will require either
populations that are reasonably closed to migration, or for the effect of
migration to be accounted for possibly either by knowledge of allele
frequencies in source populations or the identification of migrant individuals.

While it challenging to apply temporal methods to natural populations there is
a lot of promise for these approaches
\parencite{Bergland2014-ij,Machado2018-cs}. Efforts to quantify the impact of
linked selection have found obligately sexual organisms have up to an 89\%
reduction in genome-wide diversity over long time periods
\parencite{McVicker2009-ax,Elyashiv2016-vt,Corbett-Detig2015-gt,Coop2016-gx,Comeron2014-nh}
\gc{Thus linked selection makes a sizeable contribution to long-term allele
frequency change in some species,} and there is reason to be hopeful that we
could detect this from temporal data, which would help to resolve the
timescales that linked selection act over. In our reanalysis of the
\textcite{Barghi2019-qy} study, we find evidence of complex linked selection
dynamics, with selection pressures flipping over time due to either
environmental change, the breakup of epistatic combinations or advantageous
haplotypes. Such patterns would be completely obscured in samples only from
contemporary populations. \gc{Thus,} we can hope to have a much richer picture
of the\gc{ impact of selection} as temporal sequencing becomes more common,
\gc{allowing} us to observe the effects of ecological dynamics in genomic data
\parencite{Hairston2005-ga}.

>>>>

One concern is that these covariances reflect the localized impact of a few
large-effect loci rather than selection on a polygenic trait. Since our
covariances are essentially averages over loci, the covariance estimate can be
strongly affected by a single outlier region. To test whether large outlier
regions drive the genome-wide signal we see in the \textcite{Barghi2019-qy}
data, we calculate the covariances in 100kb windows along the genome (we refer
to these as windowed covariances throughout) and take the median windowed
covariance, and trimmed-mean windowed covariance, as a measure of the
genome-wide covariance robust to large-effect loci. These robust estimates
(Supplementary Table \ref{supp:table-trimmed-mean} and Supplementary Figure
\ref{suppfig:barghi-trimmed-mean}) confirm the patterns we see using the mean
covariance, confirming that genomic temporal covariances are non-zero due to
the impact of selection acting across many genomic windows.

@@

We've calculated the convergent correlation for all rows of the replicate
covariance matrices. Like temporal covariances, we visualize these through time
(Figure \ref{fig:figure-2}A), with each line representing the convergent
correlation from a particular reference generation $s$ as it varies with $t$
(shown on the x-axis). In other words, each of the colored lines corresponds to
the like-colored row of the convergence correlation matrix (upper left in
Figure \ref{fig:figure-2}A). We find these convergent covariances are
relatively weak, and decay very quickly from an initial value of about 0.1
(95\% block bootstrap confidence intervals $[0.094, 0.11]$) to around 0.01
(95\% CIs [0.0087, 0.015]) within 20 generations. This suggests that while
a reasonable fraction of the initial response is shared over the replicates,
this is followed by a rapid decay, a result consistent with the primary
finding of the original \textcite{Barghi2019-qy} study: that alternative loci
contribute to longer term adaptation across the different replicates.

@@

Next, we reanalyzed the Longshanks selection experiment, which selected for
longer tibiae length relative to body size in mice, leading to a response to
selection of about 5 standard deviations over the course of twenty generations
\parencite{Marchini2014-de,Castro2019-uk}. This study includes two independent
selection lines, Longshanks 1 and 2 (LS1 and LS2), and an unselected control
line (Ctrl).  Consequently, this selection experiment offers a useful control
to test our between-replicate covariances: we expect to see positive
between-replicate covariance in the comparison between the two Longshanks
selection lines, but not between the two pairwise comparisons between the
control line and each of the two Longshanks lines. We find that this is
the case (gray confidence intervals in Figure \ref{fig:figure-2} C), with the
two Longshanks comparisons to the control line not being significantly
different from zero, while the comparison between the two Longshanks lines is
statistically significantly different from zero (CIs [0.0129, 0.0400]).

One finding in the Longshanks study was that two major-effect loci showed
parallel frequency shifts between the two selection lines: a region harboring
the gene Nkx3-2 known to be involved in limb development, and another region
containing six other candidate genes. We were curious to what extent our
genome-wide covariances were being driven by these two outlier large-effect
loci, so we excluded them from the analysis. Since we do not know the extent to
which linkage disequilibrium around these large-effect loci affects neighboring
loci, we took the conservative precaution of excluding the entire chromosomes
these loci reside on (chromosomes 5 and 10), and re-calculating the temporal
covariances. We find excluding these large effect loci has little impact on the
confidence intervals (blue confidence intervals in Figure \ref{fig:figure-2} C),
indicating that these across-replicate covariances are indeed driven by a large
number of loci. This is consistent with a signal of selection on a polygenic
trait driving genome-wide change, although we note that large-effect loci can
contribute to the indirect change at unlinked loci
\parencite{Robertson1961-ho,Santiago1995-hx}.

The presence of an unselected control line provides an alternative way to
partition the effects of linked selection and genetic drift: we can compare the
total variance in allele frequency change of the control line (which excludes
the effect of artificial selection on allele frequencies) to the total variance
in frequency change of the Longshanks selection lines. We can partition the
allele frequency change between the two timepoints (20 generations apart) for a
Longshanks line as $\Delta p_{t,\mathrm{LS1}} = \Delta_{_{D}}
p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} + \Delta_{_S}
p_{t,\mathrm{LS}}$ where these terms are the decomposition in the allele
frequency change due to drift in Longshanks replicate 1 ($\Delta_{_D}
p_{t,\mathrm{LS1}}$), selection unique to the LS1 replicate ($\Delta_{_U}
p_{t,\mathrm{LS1}}$), and selection response shared between the two Longshanks
replicates ($\Delta_{_S} p_{t,\mathrm{LS}}$) respectively (and similarly for
the Longshanks two line, LS2). By construction we will assume that each of
these terms are uncorrelated within replicates, and that only the shared term
covaries between the replicates.  Assuming that we can approximate the
contribution of genetic drift in the LS lines as the variance in allele
frequency change observed in the control, i.e.  $\var(\Delta
p_{t,\mathrm{Ctrl}}) = \var(\Delta_{_{D}}
p_{t,\mathrm{LS2}})=\var(\Delta_{_{D}} p_{t,\mathrm{LS2}})$, then we can
estimate the increase in variance in allele frequency change due to selection
as $(\var(\Delta p_{t,\mathrm{LS1}}) + \var(\Delta p_{t,\mathrm{LS2}}))/2 -
\var(\Delta p_{t,\mathrm{Ctrl}})$ and the shared effect of selection across
selected lines as $\cov(\Delta p_{t,\mathrm{LS1}}, \Delta p_{t,\mathrm{LS2}})$
(see Supplementary Material Section \ref{supp:replicate-g} for more details).
We estimate at least 32\% (95\% CI $[21\%, 48\%]$) of the variance in allele
frequency change is driven by the effects of selection, of which 14\% (95\% CI
$[3\%, 33\%]$) is estimated to be unique to a selection line, and 17\% (95\% CI
$[9\%, 23\%]$) is the effect of shared selection between the two Longshanks
selection lines (and the value of the convergence correlation between the
Longshanks lines, a related statistic, is 0.18, 95\% CI [0.0743, 0.254]).

@@

Finally, we observed that in the longest study  we analyzed
\parencite{Barghi2019-qy}, some genome-wide temporal covariances become
negative at future timepoints (see the first two rows in Figure
\ref{fig:figure-1} A). This shows that alleles that were on average going up
initially are later going down in frequency, i.e. that the average direction of
selection experienced by alleles has flipped. This must reflect either a change
in the environment or the genetic background, due to epistatic relationships
among alleles altered by frequency changes or recombination breaking up
selective alleles.  Such reversals of selective dynamics could be occurring at
other timepoints but the signal of a change in the direction of selection at
particular loci may be washed out when we calculate our genome-wide average
temporal covariances. To address this limitation, we calculated the
distribution of the temporal covariances over 100kb windowed covariances
(Figure \ref{fig:figure-3} shows these distributions pooling across all
replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance estimate of each genomic window will be noisy, due to sampling and
genetic drift, and the neutral distribution of the covariance is complicated
due to linkage disequilibria (which can occur over long physical distances in
E\&R and selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}). To
address this, we have developed a permutation-based procedure that constructs
an empirical null distribution by randomly flipping the signs of the allele
frequency changes per-genomic window. This destroys the systematic covariances
created by linked selection and creates a sampling distribution of the
covariances spuriously created by neutral genetic drift while preserving the
complex dependencies between adjacent loci created by linkage disequilibrium.
This empirical neutral null distribution is conservative in the sense that the
variances of the covariances are wider than expected under drift alone as they
include the effect of selection on the allele frequency change within a
time-interval, just not between time-intervals. We see (Figure
\ref{fig:figure-3} A and B) that windowed temporal covariances between close
timepoints are skewed positive (a heavy right tail), while between more distant
timepoints these windowed temporal covariances tend to shift to become more
negative (a heavy left tail).  We quantified the degree to which the left and
right tails are inflated compared to the null distribution as a function of
time, and see excesses in both tails in Figure \ref{fig:figure-3} C. This
finding is also robust to sign-permuting allele frequency changes on a
chromosome-level, the longest extent that gametic linkage disequilibria can
extend (Supplementary Figure \ref{suppfig:barghi-tailprobs-seqid}). We see a
striking pattern that the windowed covariances not only decay towards zero, but
in fact become negative through time, consistent with many regions in the
genome having had a reversed fitness effect at later timepoints.

@@

These estimates of the contribution of selection could be refined by using
patterns of LD and recombination which would allow us to more fully
parameterize a linked-selection model of temporal allele frequency change
\parencite{Buffalo2019-io}. The basic prediction is that regions of higher
linkage disequilibrium and lower recombination should have greater temporal
autocovariance than regions with lower LD and higher recombination. However,
one limitation of these pooled sequence datasets is that none of the studies we
reanalyzed estimated linkage disequilibria data for the evolved populations.
While there are LD data for a natural population of \emph{D. simulans}
\parencite{Signor2018-wg,Howie2018-ay},  we did not find a relationship between
temporal covariance and LD.  We believe this is driven by the idiosyncratic
nature of LD in evolve-and-resequence populations, which often extends over
large genomic distances \parencite{Nuzhdin2013-gf,Kelly2019-dc}. Future studies
complete with LD data and recombination maps would allow one to disentangle the
influence of closely linked sites from more distant sites in causing temporal
autocovariance, and allow the fitting of more parametric models to estimate
population parameters such as the additive genetic variation for fitness
directly from temporal genomic data alone \parencite{Buffalo2019-io}.

Our primary focus here has been on evolution in laboratory populations. It is
unclear whether we should expect a similar impact of selection in natural
populations. In some of these experiments, selection pressures may have been
stronger or more sustained that in natural populations
\parencite{Hendry1999-zu,Hairston2005-ga}. Conversely, these lab
populations were maintained at very small effective population sizes, estimated
at 300, 450, and 45 for the \textcite{Barghi2019-qy}, \textcite{Kelly2019-dc},
and \textcite{Castro2019-uk} studies respectively, which will amplify the role
of genetic drift. The advantage of lab experiments is that they are closed
populations, in natural populations temporal covariance could also arise from
the systematic migration of alleles from differentiated populations.
Adapting these methods to natural populations will require either
populations that are reasonably closed to migration, or for the effect of
migration to be accounted for possibly either by knowledge of allele
frequencies in source populations or the identification of migrant individuals.

While it challenging to apply temporal methods to natural populations there is
a lot of promise for these approaches
\parencite{Bergland2014-ij,Machado2018-cs}. Efforts to quantify the impact of
linked selection have found obligately sexual organisms have up to an 89\%
reduction in genome-wide diversity over long time periods
\parencite{McVicker2009-ax,Elyashiv2016-vt,Corbett-Detig2015-gt,Coop2016-gx,Comeron2014-nh}
Thus linked selection makes a sizeable contribution to long-term allele
frequency change in some species, and there is reason to be hopeful that we
could detect this from temporal data, which would help to resolve the
timescales that linked selection act over. In our reanalysis of the
\textcite{Barghi2019-qy} study, we find evidence of complex linked selection
dynamics, with selection pressures flipping over time due to either
environmental change, the breakup of epistatic combinations or advantageous
haplotypes. Such patterns would be completely obscured in samples only from
contemporary populations. Thus, we can hope to have a much richer picture
of the impact of selection as temporal sequencing becomes more common,
allowing us to observe the effects of ecological dynamics in genomic data
\parencite{Hairston2005-ga}.
